Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd
Revision in intro.

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite universality of the phenomenon of leadership, the function leaders serve across populations and social contexts are highly diverse and expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

Theoretical literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often take on essential roles in within-group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group projects, and administering sanctions to individuals who fail to appropriately contribute or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies, representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions of leaders.

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Increased sedentarization and defensibility of resources is often associated with both social stratification and broader institutionalized managerial roles including stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources. Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources [@lopez_evolutionary_2017].

@@

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, many managerial scholars have adopted comparative approaches to better understand cultural or cross-national diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, @house2013strategic suggest leaders moderate their behavior through culturally learned values and leadership-specific values.

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to influence group processes can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formal institutionalized roles and informal ephemeral roles, are thought to be commonly associated with high social status (as previously discussed). Leadership roles are also widely associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many WEIRD industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also generally viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].

@@

```{r fivefold, fig.width=10, fig.cap="The intersections of evidence for our five basic elements of leadership among all text records (an 'upset' plot) [@gehlenborg2019]. Horizontal bars (left): The total number of text records in each element. Dots (bottom): Elements that both have evidence (intersection). Vertical bars (top): The number of text records in that intersection. Red: The number of text records involving shamans or leaders with supernatural qualities. "}

>>>>

## The functions of leadership across cultures and contexts
<!-- INTRO -->
Despite the universality of leadership, the functions leaders serve across populations and social contexts are highly diverse and expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when group demands require a managerial presence [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993].

Theoretical literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often manage within-group coordination and collective action, including conflict resolution within or between households or kin groups, managing individual investments to group projects, and administering sanctions when individuals fail to appropriately contribute or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also common across societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of smaller-scale community leaders. Within communities leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are suggested to be fairly widespread functions of leaders.

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Increased sedentarization and defensibility of resources is often associated with social stratification, broader institutionalized managerial roles, and stronger authority of community leaders and leadership beyond the community level [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985; @panchanathan_indirect_2004]. Among horticulturalists and agriculturalists managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of critical resources creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military [@lopez_evolutionary_2017].

@@

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, some managerial scholars have adopted comparative methods. @behrendt_integrative_2017 suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. 
<!-- More integrated models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning.  -->
Culturally Endorsed Implicit Leadership Theory (CLT), suggests social norms impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on and hundreds of publications. *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* [@house2013strategic] uses data from over 1,000 CEOs and over 5,000 senior executives from varied corporations across 24 countries to address (1) national cultural influences on the behaviors of leaders, (2) the behaviors of CEOs typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. The study concluded cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, @house2013strategic suggest leaders moderate their behavior through culturally learned values and leadership-specific values.

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to influence group processes can provide enormous benefits to leaders, even if 'just for a while.' Leaders in any context can promote policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formal and informal roles, are commonly associated with high social status and numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. Leaders and high-status individuals are more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many WEIRD organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also generally viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].

@@

```{r fivefold, fig.width=10, fig.cap="The intersections of evidence for our five basic elements of leadership among all text records (an 'upset' plot) [@gehlenborg2019]. Horizontal bars (left): The total number of text records in each element. Dots (bottom): Elements that both have evidence (intersection). Vertical bars (top): The number of text records in that intersection (with lower limit set at 3). Red: The number of text records involving shamans or leaders with supernatural qualities. "}
